Models

### 2021-04-06 analysis/05_model_metrics.Rmd
incorporate null model samples

metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  mutate(logvalue = log(value + abs(floor(min(na.omit(value))))),
         year_scaled = scale(year)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
  filter(!is.na(value)) 
```
####Data exploration

@@

metrics %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  ggplot() +
  geom_histogram(aes(x = value),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of raw values")
```

@@

```{r}
# model for overall trend (random effect w/varying intercept)
metric_mean_fit <- metrics %>%
  filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S_int")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (1|study_id/rarefyID), data = .x,
                  control = lmerControl(optCtrl = list(maxit = 1e9,
                                                    maxfun = 1e9,
                                                    xtol_abs = 1e-11,
                                                    ftol_abs = 1e-11,
                                                    maxeval = 1e9)))),
         null_model = map(data, ~lmer(logvalue ~ (1|study_id/rarefyID), data = .x,
                  control = lmerControl(optCtrl = list(maxit = 1e9,
                                                    maxfun = 1e9,
                                                    xtol_abs = 1e-11,
                                                    ftol_abs = 1e-11,
                                                    maxeval = 1e9)))),

           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

metric_mean_fit_slope <- metrics %>%
  filter(metric %in%  c("FRic", "FEve", "FDiv", "S_int")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

model_comp <- metric_mean_fit %>%
  select(metric, no_slope_estimate = estimate, no_slope_pval = p.value) %>%
  left_join(metric_mean_fit_slope %>% 
  select(metric, slope_estimate = estimate, slope_pval = p.value ))
```

@@

```{r}

>>>>

metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
  # pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  filter(!is.na(value)) %>%
  mutate(logvalue = log(value + 1),
         year_scaled = scale(year)) # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values

```
####Data exploration

@@

metrics %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic")) %>%
  ggplot() +
  geom_histogram(aes(x = value),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of raw values")
```

@@

```{r}
#find rarefyID's that have inf value for some metrics and exclude
cont_inf <- metrics %>% filter(is.infinite(value), !metric %in% c("qual.FRic",  "SES_qual.FRic")) %>% pull(rarefyID)

#get rarefyID's that don't have enough null samples
missing_null <- bind_rows(metrics %>% 
  filter(n_missing_nulls == 1, commplete_null_samps == FALSE),
  metrics %>% filter(n_missing_nulls > 100)) %>%
  pull(rarefyID) %>%
  unique()

model_data <- metrics %>% 
  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
  mutate(study_id = substr(rarefyID, 1, 3))

# model for overall trend (random effect w/varying intercept)
# metric_mean_fit <- model_data %>%
#   filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
#   group_nest(metric) %>%
#   mutate(model = map(data, ~lmer(log(value + 1) ~ year_scaled + (1|study_id/rarefyID), data = .x, REML = FALSE,
#                   control = lmerControl(optCtrl = list(maxit = 1e9,
#                                                     maxfun = 1e9,
#                                                     xtol_abs = 1e-11,
#                                                     ftol_abs = 1e-11,
#                                                     maxeval = 1e9)))),
#          null_model = map(data, ~lmer(logvalue ~ (1|study_id/rarefyID), data = .x, REML = FALSE,
#                   control = lmerControl(optCtrl = list(maxit = 1e9,
#                                                     maxfun = 1e9,
#                                                     xtol_abs = 1e-11,
#                                                     ftol_abs = 1e-11,
#                                                     maxeval = 1e9)))),
#            
#            test = map2(model, null_model, anova),
#            test_res = map(test, tidy),
#            coef = map(model, tidy)
#            ) %>% 
#   unnest(cols = c(test_res)) %>%
#   select(metric, model, coef, p.value) %>%
#   filter(!is.na(p.value)) %>%
#   unnest(cols = c(coef)) %>%
#   filter(term == "year_scaled") %>%
#   select(-c(effect, group, term, statistic))

SES_rand_slope <- model_data %>%
  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

metric_rand_slope <- model_data %>%
  filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = FALSE)),
           null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = FALSE)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

# model_comp <- metric_mean_fit %>%
#   select(metric, no_slope_estimate = estimate, no_slope_pval = p.value) %>%
#   left_join(metric_mean_fit_slope %>% 
#   select(metric, slope_estimate = estimate, slope_pval = p.value ))

metric_model_table <- bind_rows(select(SES_rand_slope, -model), select(metric_rand_slope, -model))
```

Kitchen sink model
```{r}
all_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis", "S")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + realm + (year_scaled|study_id/rarefyID), 
                                 data = .x)))#,
  #          null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
  #                                data = .x)),
  #          test = map2(model, null_model, anova),
  #          test_res = map(test, tidy),
  #          coef = map(model, tidy)
  #          ) %>% 
  # unnest(cols = c(test_res)) %>%
  # select(metric, model, coef, p.value) %>%
  # filter(!is.na(p.value)) %>%
  # unnest(cols = c(coef)) %>%
  # filter(term == "year_scaled") %>%
  # select(-c(effect, group, term, statistic))

@@

Get study level slop estimates
```{r}
get_blup <- function(metric, model){
  coef(model)$study_id %>%
    rownames_to_column("study_id") %>%
    mutate(metric = metric)
}

blup <- pmap_dfr(bind_rows(select(SES_rand_slope, metric, model), select(metric_rand_slope,  metric, model)), get_blup) %>%
  select(-"(Intercept)") %>%
  pivot_wider(names_from = "metric", values_from = "year_scaled")

#pairwise plots using GGally
ggpairs(blup %>% 
          select(-c("study_id", "FDis", "FDiv", "FEve", "FRic")))

#pairwise plots using GGPlot
gatherpairs <- function(data, ..., 
                        xkey = '.xkey', xvalue = '.xvalue',
                        ykey = '.ykey', yvalue = '.yvalue',
                        na.rm = FALSE, convert = FALSE, factor_key = FALSE) {
  vars <- quos(...)
  xkey <- enquo(xkey)
  xvalue <- enquo(xvalue)
  ykey <- enquo(ykey)
  yvalue <- enquo(yvalue)

  data %>% {
    cbind(gather(., key = !!xkey, value = !!xvalue, !!!vars,
                 na.rm = na.rm, convert = convert, factor_key = factor_key),
          select(., !!!vars)) 
  } %>% gather(., key = !!ykey, value = !!yvalue, !!!vars,
               na.rm = na.rm, convert = convert, factor_key = factor_key)
}

blup %>% 
  gatherpairs("SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
  ggplot(., aes(x = .xvalue, y = .yvalue)) +
      geom_point() + 
      #geom_smooth(method = 'lm') +
      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
      scale_color_brewer(type = 'qual')
  }

blup %>%
  select(-c("FDis", "FDiv", "FEve", "FRic")) %>%
  pivot_longer(-one_of("study_id", "S"), names_to = "metric") %>%
  ggplot(aes(x = S, y = value)) +
  geom_point() +
  facet_wrap(~metric) +
  xlim(-1.3, 1.3) +
  ylim(-0.6, 0.6) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) + 
  ylab("Metric Slope") +
  xlab("Richness Slope")

blup %>%
  select(-c("FDis", "FDiv", "FEve", "FRic")) %>%
  pivot_longer(-one_of("study_id", "SES_FRic"), names_to = "metric") %>%
  ggplot(aes(x = SES_FRic, y = value)) +
  geom_point() +
  facet_wrap(~metric) +
  xlim(-.25, .25) +
  ylim(-0.6, 0.6) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) + 
  ylab("Metric Slope") +
  xlab("Functional Richness Slope")
```
